Automating Political Bias Prediction

### 2016-04-03 manuscript/political_bias_prediction.tex
Polished text.

% the name(s) of the author(s) follow(s) next
%
% NB: Chinese authors should write their first names(s) in front of
% their surnames. This ensures that the names appear correctly in
% the running heads and the author index.
%
\author{
%Felix Bie\ss{}mann%
\thanks{}
%
\authorrunning{Political Bias Prediction}}
% (feature abused for this document to repeat the title also on left hand pages)

@@

\begin{abstract} 
Every day media generate large amounts of text. An unbiased view on what media report on requires an understanding of the political bias of media content. Assistive technology for estimating the political bias of texts can be helpful in this context. This study proposes a simple statistical learning approach to predict political bias from text. Standard text features extracted from speeches and manifestos of political parties are used to predict political bias in terms of political party affiliation and in terms of political views. Results indicate that political bias can be predicted with well above chance accuracy. Mistakes of the model can be interpreted with respect to changes of policies of political actors. Two approaches are presented to make the results more interpretable: a) discriminative text features are related to the political orientation of a party and b) sentiment features of texts are correlated with a measure of political power. Political power appears to be strongly correlated with positive sentiment of a text. To highlight some potential use cases a web application shows how the model can be used for texts for which the political bias is not clear such as news articles.
\end{abstract}

@@

Automated predictions of political bias can be problematic for ethical reasons. One trusts political experts, as they will take responsibility for what they say about a text, and they can explain their decisions. This is a key difference to many statistical learning approaches. Not only is the responsibility question not entirely clear, it can also be difficult to interpret some of the decisions. In order to validate the predictions of the models we devise two strategies that allow for better interpretations of the models. First we are looking at univariate measures of how discriminative the features are that the model uses. This somewhat allows to relate the political texts to the kind of information that political experts use, too. Second we use sentiment analysis to investigate whether this aspect of language also has discriminatory power. \\

In the following \autoref{sec:data} gives an overview of data acquisition and preprocessing, \autoref{sec:model} presents the model, in \autoref{sec:results} the results are discussed and \autoref{sec:conclusion} concludes with some interpretations of the results and future research directions.

@@

\paragraph{Parliament discussion data} Parliament texts are annotated with the respective party label, which we take here as a proxy of political bias. The texts of parliament protocols are available through the website of the german bundestag\footnote{\url{https://www.bundestag.de/protokolle}}, we used an open source API to that data to query the data in cleaned and structured format\footnote{\url{https://github.com/bundestag}}. This API returns for each speech the speaker and its party affiliation. In total we obtained 22784 speeches for the 17th legislative period and 11317 speeches for the 18th period, queried until March 2016.

\paragraph{Party manifesto data}
For party manifestos we also use an openly accessible API provided by the Wissenschaftszentrum Berlin (WZB). The API is released as part of the {\em Manifestoproject} \cite{manifesto}. The data released in this project comprises the complete manifestos for each party that ran for election enriched with annotations by political experts. Each sentence (in some cases also parts of sentences) are labels with one of 56 political labels. Examples of these labels are {\em pro/contra protectionism, decentralism, centralism, pro/contra welfare, ...}. The set of labels was developed by political scientists at the WZB and released for public use. We obtained all manifestos of parties that were running for election in this and the last legislative period. In total this resulted in 29451 political statements that had two types of labels, for one we stored the party affiliation of each political statement. This party affiliation label was used to evaluate the party evaluation classifiers trained on the parliament speeches. For this purpose we constrained the data to only those parties that were elected into the parliament. Next to the party affiliation we were mainly interested in the political view labels annotated by human experts. For the analyses based on political view labels we considered all parties, also those that did not make it into the parliament. \\
The length of each annotated statement in the party manifestos was rather short. The longest statement was 522 characters long, the 25\%/50\%/75\% percentiles were 63/95/135 characters, measured in words the longest data point was 65 words and the 25\%/50\%/75\% percentiles were 8/12/17 words, respectively. This can be considered as a very valuable property of the data set, because it allows a fine grained resolution of party manifestos. However for a classifier (as well as for humans) such short sentences can be rather difficult to classify. In order to obtain less 'noisy' data points from each party all statements were aggregated into political topics using the manifesto code labels: each political view label is a three digit code, the first digit represents the political domain. In total there were eight political domains (topics): {\em External Relations, Freedom and Democracy, Political System, Economy, Welfare and Quality of Life, Fabric of Society, Social Groups} and a topic {\em undefined}, for a complete list see also \cite{leftright}. These 8 topics were used to aggregate all statements in each manifesto into topics. Most party manifestos covered all eight of them, some party manifestos in the 17th Bundestag only covered seven.

\subsection{Bag-of-Words Vectorization}\label{sec:bow-vectorization}
First each data set was segmented into semantic units; in the case of parliament discussions this were the speeches, in the case of the party manifesto data semantic units were the sentences or sentence parts associated with one of the 56 political view labels.  Parliament speeches were often interrupted; in this case each uninterrupted part of a speech was considered a semantic unit. Strings of each semantic unit were tokenized and transformed into bag-of-word vectors as implemented in scikit-learn \cite{scikit-learn}. The general idea of bag-of-words vectors is to simply count occurences of words (or word sequences, also called {\em n-grams}) for each data point. A data point is usually a documents, here it is the semantic units of parliament speeches and manifesto sentences, respectively. The text of each semantic unit is transformed into a vector $\vec{x}\in\mathds{R}^d$ where $d$ is the size of our dictionary; the $w$th entry of $\vec{x}$ contains the (normalized) count of the $w$th word (or sequence of words) in our dictionary. We tried several options for vectorizing the speeches, including term-frequency-inverse-document-frequency normalisation using n-gram patterns up to size $n=3$ and several cutoffs for discarding too frequent and too infrequent words. All of these hyperparameters were subjected to hyperparameter optimization as explained in \autoref{sec:crossvalidation}.

\section{Classification Model and Training Procedure}\label{sec:model}
We used a multinomial logistic regression model in order to classify bag-of-words feature vectors. Let $y\in\{1,2,\dots,K\}$ be the true  label (party affiliation or political view, see \autoref{sec:data}), where $K$ is the total number of labels and $\vec{W}=[\vec{w}_1,\dots,\vec{w}_K]\in\R^{d\times K}$ is the concatenation of the weight vectors $\vec{w}_k$ associated with the $k$th party then 
\begin{eqnarray}\label{eq:logreg_multiclass}
p(y=k|\vec{x},\vec{W}) = &\frac{e^{z_k}}{\sum_{j=1}^K e^{z_j}} \qquad \textrm{with }  z_k=&\vec{w}_k^{\top}\vec{x} \\\nonumber
\end{eqnarray}
%
We estimated $\vec{W}$ using quasi-newton gradient descent. The optimization function was obtained by adding a penalization term to the negative log-likelihood of the multinomial logistic regression objective and the optimization hence found the $\vec{W}$ that minimized
\begin{equation}\label{eq:objective}
L(\vec{W}, \vec{x}, \gamma) = - \log{\frac{e^{z_k}}{\sum_{j=1}^K e^{z_j}}}+ \gamma \| \vec{W} \|_{F}
\end{equation}
Where $\|~\|_F$ denotes the Frobenius Norm and $\gamma$ is a regularization parameter controlling the complexity of the model. 
 We optimized the regularization parameter on a log-scaled grid from $10^{-4,\dots,4}$ and the regularization constant was adopted to reflect asymmetric class frequency distributions. The performance of the model was optimized using the classification accuracy, but we also report all other standard measures, precision ($TP / (FP + TP$), recall ($TP / (TP + FN)$) and f1-score ($2\times (Prec. \times Rec) / (Prec + Rec.)$).

We considered three different classification problems: 
\begin{enumerate}
\item {\bf Classification of party affiliation} 
\item {\bf Classification of government membership}
\item {\bf Classification of political views}
\end{enumerate}

Party affiliation is a five class problem for the 17th legislation period, and a four class problem for the 18th legislation period. Government membership is a binary problem. Political view classification is based on the labels of the manifestoproject, see also \autoref{sec:data}, and is a 56 class problem. 
For each of first two problems, party affiliation and government membership prediction, we trained classifiers on the parliament speeches. For the third problem we trained only on the manifesto data as only for those the political view labels were available.

\subsection{Optimisation of Model Parameters}\label{sec:crossvalidation}
The model pipeline contained a number of {\em hyperparameters} that we did not want to tune by hand. 
We first split the training data into a training data set that was used for optimization of hyperparameters and an held-out test data set for evaluating how well the model performs on in-domain data; next to the in-domain test data, we also used out-of domain data wherever possible in order to evaluate how well the model would generalize to text domains that the model has not seen during training. The test data for hyperparameter optimization was 10\% of the entire training data. Hyperparameters were optimized using grid search and 3-fold cross-validation within the training set only: Within the training set a cross-validation split was made to obtain train/test data for the grid search and for each setting of hyperparameters the entire pipeline was trained and evaluated -- no data from the in-domain evaluation data or the out-of-domain evaluation data were used for hyperparameter optimization. For the best setting of all hyperparameters, we trained the pipeline again on all training data and evaluated it on both evaluation data sets. For party affiliation prediction and government membership prediction the training and test set were 90\% and 10\%, respectively, of all data in a legislative period. Out-of-domain evaluation data were the texts from party manifestos. For the political view prediction setting there was no out-of-domain evaluation data, so all labeled manifesto sentences in both legislative periods were split into a training and evaluation set of 90\% (train) and 10\% (evaluation).

\subsection{Sentiment analysis}\label{sec:sentiment_analysis_methods}
We used a publicly available key word list to extract sentiments from the text subjected to classification \cite{remquahey2010}. The sentiment index used for attributing positive or negative sentiment to a text was computed  as the cosine similarity between bag-of-word vectors $\vec{x}\in\R^d$ and sentiment vector $\vec{s}\in\R^d$ which was constructed by simply taking the values from the sentiment dictionary.

\subsection{Analysing bag-of-words features}\label{sec:correlations_methods}
In order to get  better picture of which features contribute to the classification we computed correlation coefficients between each word and each party. Often linear models are interpreted by just looking at the weights of the model. Since the text features are obtained from real data and thus correlated we cannot inspect the model by interpreting the coefficients of $\vec{W}$, see e.g. \cite{Haufe2013}.  We thus only use univariate correlation to investigate how discriminative features are. The words corresponding to the top positive and negative correlations are shown in section \autoref{sec:word_party_correlations}.

@@

The following sections provide an overview of the results for both types of political bias prediction tasks, party affiliation and political view. Some interpretations of discriminative features are given one section will highlight a finding on the correlation of political power and positive sentiment of political texts.

\subsection{Predicting political party affiliation}
The results for the political party affiliation prediction on held-out parliament data and on evaluation data are shown in \autoref{tab:results_17} for the 17th Bundestag and in \autoref{tab:results_18} for the 18th Bundestag, respectively. 
Shown are the evaluation results for in-domain data (held-out parliament speech texts) as well as the out-of-domain data results obtained from testing the models on the texts extracted from party manifestos. When looking at predicting party affiliation from text data from the same domain that was used for training precision and recall values of above 0.6 are obtained. \\

For out-of domain data the models yield significantly lower precision and recall values between 0.3 and 0.4.  
A main factor that made the prediction on the out-of-domain prediction task particularly difficult is the short length of the strings to be classified, see also \autoref{sec:data}. In order to investigate whether this low out-of-domain prediction performance was due the domain difference (parliament speech vs manifesto data) or due to the short length of the data points, we aggregated the manifesto data based on the topic. The manifesto codes contain labels for different topics, which were used to concatenate the data points of each party to only eight data points per party, each data point corresponding to one of the eight topics defined and annotated by the manifesto project annotators. The results are shown in table \autoref{tab:results_17_topic} and \autoref{tab:results_18_topic} and clearly show that when the texts to be classified and the word count statistics are sufficiently dense the classification performance on out of domain data becomes close to precisions of 1.0.

@@

\begin{table}[t]
\begin{center}
\begin{tabular}{lcccc}
& \multicolumn{4}{c}{\bf 17th Bundestag} \\
    &         precision    &recall &  f1-score  & N  \\
    \hline
cducsu     &  0.64  &    1.00  &    0.78    &     7\\
       fdp    &   1.00    &  1.00    &  1.00    &     7\\
    gruene  &     1.00  &    0.86  &    0.92    &     7\\
     linke    &   1.00   &   1.00     & 1.00    &     7\\
       spd   &    0.80   &   0.50    &  0.62     &    8\\
    \hline
    \hline
    avg / total  &     0.88   &   0.86   &   0.86  &      36\\
\end{tabular}
\quad
\begin{tabular}{lcccc}
& \multicolumn{4}{c}{\bf 18th Bundestag} \\
    &         precision    &recall &  f1-score  & N  \\
    \hline
  cducsu     &  0.50   &   1.00  &    0.67     &    8\\
       gruene  &     0.00   &   0.00   &   0.00  &       8\\
      linke     &  1.00   &   0.88   &   0.93  &       8\\
        spd     &  0.56  &    0.62  &    0.59   &      8\\

avg / total   &    0.51 &     0.62   &   0.55   &     32\\
\end{tabular}
\end{center}
\caption{
\label{tab:results_topic}
{\bf Topic level classification performance} on the party affiliation prediction problem for data from the evaluation set (manifesto texts) of the 17th legislative period. Instead of sentence level predictions as in \autoref{fig:results_17}, here predictions per political topic are shown (see \autoref{sec:data}); in contrast to single sentence level predictions (see \autoref{tab:confusion}) the predictions made on topic level are in many cases reliable. One effect present in the single sentence level predicitions however is even more pronounced in the topic level predictions: the policy of the green party manifesto is completely absorbed in the parliament speeches of the right-conservative CDU/CSU party -- all manifesto topics of the green party are predicted to be from one of the parties of the governing coalition, CDU/CSU or SPD.}
\end{table}

%trained with cv on bundestag speeches 17
%tested on topic aggregated (first digit of manifestocode) manifesto data
%
%cducsu       0.64      1.00      0.78         7
%        fdp       1.00      1.00      1.00         7
%     gruene       1.00      0.86      0.92         7
%      linke       1.00      1.00      1.00         7
%        spd       0.80      0.50      0.62         8
%
%avg / total       0.88      0.86      0.86        36
%
%Confusion Matrix (rows=true, cols=predicted)
%cducsu, fdp, gruene, linke, spd
%[7, 0, 0, 0, 0]
%[0, 7, 0, 0, 0]
%[0, 0, 6, 0, 1]
%[0, 0, 0, 7, 0]
%[4, 0, 0, 0, 4]
%
%0.86111111111111116
%
%trained with cv on bundestag speeches 18
%tested on topic aggregated (first digit of manifestocode) manifesto data
%
%*** Training Set (CVd) ***
%             precision    recall  f1-score   support
%
%     cducsu       0.50      1.00      0.67         8
%     gruene       0.00      0.00      0.00         8
%      linke       1.00      0.88      0.93         8
%        spd       0.56      0.62      0.59         8
%
%avg / total       0.51      0.62      0.55        32
%
%Confusion Matrix (rows=true, cols=predicted)
%cducsu, fdp, gruene, linke, spd
%[8, 0, 0, 0]
%[4, 0, 0, 4]
%[1, 0, 7, 0]
%[3, 0, 0, 5]
%

In order to investigate the errors the models made we extracted confusion matrices for the predictions on the out-of-domain evaluation data, the manifesto data. So the model, trained on party speeches in the parliament was tested on the texts parties published before the elections in order to get elected. One example in \autoref{tab:confusion} shows that these mistakes the model makes can give valuable insights into the changes in policy a party decides to make. The green party was promoting policies for renewable energy and agains nuclear energy in their manifestos prior to both legislative periods. When training a classifier on what representatives in the parliament say however, the statements of the green party are more often predicted to be from the conservative government than from the party that originally promoted these green ideas. This effect is also reflected in the precision values on out-of-domain data: The green party has the highest precision values by a large margin in both legislation periods, indicating that there is a relatively high true-positive rate. A similar effect was can be observed with the social democratic party (SPD): Evaluating the parliament classifier on manifestos it seems that a significant amount of speeches get classified as conservative government while being from the SPD. \\

Next to the party affiliation we also predicted whether or not a text is from a party that belonged to a given governing coalition of the Bundestag. In \autoref{tab:results_binary_17} and \autoref{tab:results_binary_18} the results are shown for the 17th and the 18th Bundestag, respectively. While the in-domain evaluation precision and recall values are much better in this case, reaching values close to 0.9, the out-of-domain evaluation drops again to values between 0.6 and 0.7.

% Confusion matrix sentence level
\begin{table}[t]\label{tab:conf_mat_four_class}
\begin{tabular}{lcc|ccccc}
 \multicolumn{8}{c}{\bf 17th Bundestag}\\
 \\
&&& \multicolumn{5}{c}{Predicted}\\
&&& cducsu & fdp& gruene& linke& spd\\
\hline
\multirow{5}{*}{\rotatebox{90}{\pbox{3cm}{\centering True}}}& &cducsu &1186 &289& 178& 198& 179\\
&&fdp &882& 658& 236& 329& 214\\
&&gruene &1174& 404& 764& 941& 464\\
&&linke &388& 92& 214& 806& 201\\
&&spd &999& 268& 240& 398& 373\\
\end{tabular}
\hfill
\begin{tabular}{lcc|cccc}
 \multicolumn{7}{c}{\bf 18th Bundestag}\\
 \vspace{1em}\\
&&& \multicolumn{4}{c}{Predicted}\\
&&& cducsu & gruene& linke& spd\\
\hline
\multirow{4}{*}{\rotatebox{90}{\pbox{4.7cm}{\centering True}}}&&cducsu&1912& 156& 331& 584\\
&&gruene&2092& 827& 1311& 1444\\
&&linke&596& 186& 1216& 557\\
&&spd&1284& 226& 563& 916\\
\end{tabular}
\vspace{1em}
\caption{\label{tab:confusion} {\bf Confusion matrices (sentence level)} for predictions on evaluation data (party manifestos) for classifiers trained on parliament speeches for the 17th legislative period (left) and 18th legislative period (right); the most prominent effect is the high likelihood for a party to be taken as the strongest, governing party, cdu/csu. In particular the green party suffers from this effect: when trained on the speeches in the 18th Bundestag, 2092 statements out of the green party's manifesto were classified as belonging to the cducsu party. This can be interpreted as a change in policies of the conservative party cdu/csu towards the policies of the green party.}
\end{table}

>>>>

% the name(s) of the author(s) follow(s) next
%
% NB: Chinese authors should write their first names(s) in front of
% their surnames. This ensures that the names appear correctly in
% the running heads and the author index.
%
\author{
Felix Biessmann
\thanks{}
%
\authorrunning{Political Bias Prediction}}
% (feature abused for this document to repeat the title also on left hand pages)

@@

\begin{abstract} 
Every day media generate large amounts of text. An unbiased view on what media report on requires an understanding of the political bias of media content. Assistive technology for estimating the political bias of texts can be helpful in this context. This study proposes a simple statistical learning approach to predict political bias from text. Standard text features extracted from speeches and manifestos of political parties are used to predict political bias in terms of political party affiliation and in terms of political views. Results indicate that political bias can be predicted with above chance accuracy. Mistakes of the model can be interpreted with respect to changes of policies of political actors. Two approaches are presented to make the results more interpretable: a) discriminative text features are related to the political orientation of a party and b) sentiment features of texts are correlated with a measure of political power. Political power appears to be strongly correlated with positive sentiment of a text. To highlight some potential use cases a web application shows how the model can be used for texts for which the political bias is not clear such as news articles.
\end{abstract}

@@

When human experts determine political bias of texts they will take responsibility for what they say about a text, and they can explain their decisions. This is a key difference to many statistical learning approaches. Not only is the responsibility question not entirely clear, it can also be difficult to interpret some of the decisions. In order to validate the predictions of the models three strategies that allow for better interpretations of the models are proposed. First interpretations of the model misclassifications are related to changes in party policies. Second univariate measures of correlation between text features and party affiliation allow to relate the predictions to the kind of information that political experts use for interpreting texts. Third sentiment analysis is used to investigate whether this aspect of language has discriminatory power. \\

In the following \autoref{sec:data} gives an overview of data acquisition and preprocessing, \autoref{sec:model} presents the model, training and evaluation procedures; in \autoref{sec:results} the results are discussed and \autoref{sec:conclusion} concludes with some interpretations of the results and future research directions.

@@

\paragraph{Parliament discussion data} Parliament texts are annotated with the respective party label, which we take here as a proxy for political bias. The texts of parliament protocols are available through the website of the german bundestag\footnote{\url{https://www.bundestag.de/protokolle}}, we used an open source API to query the data in a cleaned and structured format\footnote{\url{https://github.com/bundestag}}. In total we obtained 22784 speeches for the 17th legislative period and 11317 speeches for the 18th period, queried until March 2016.

\paragraph{Party manifesto data}
For party manifestos we also use an openly accessible API provided by the Wissenschaftszentrum Berlin (WZB). The API is released as part of the {\em Manifestoproject} \cite{manifesto}. The data released in this project comprises the complete manifestos for each party that ran for election enriched with annotations by political experts. Each sentence (in some cases also parts of sentences) is annotated with one of 56 political labels. Examples of these labels are {\em pro/contra protectionism, decentralism, centralism, pro/contra welfare, ...}, for a complete list and detailed explanations on how the annotators were instructed see \cite{leftright}. The set of labels was developed by political scientists at the WZB and released for public use. We obtained all manifestos of parties that were running for election in this and the last legislative period. In total this resulted in 29451 political statements that had two types of labels: First the party affiliation of each political statement which was used to evaluate the party evaluation classifiers trained on the parliament speeches. For this purpose we constrained the data to only those parties that were elected into the parliament. Next to the party affiliation the political view labels were extracted. For the analyses based on political view labels we considered all parties, also those that did not make it into the parliament.

The length of each annotated statement in the party manifestos was rather short. The longest statement was 522 characters long, the 25\%/50\%/75\% percentiles were 63/95/135 characters, measured in words the longest data point was 65 words and the 25\%/50\%/75\% percentiles were 8/12/17 words, respectively. This can be considered as a very valuable property of the data set, because it allows a fine grained resolution of party manifestos. However for a classifier (as well as for humans) such short sentences can be rather difficult to classify. In order to obtain less 'noisy' data points from each party all statements were aggregated into political topics using the manifesto code labels. Each political view label is a three digit code, the first digit represents the political domain. In total there were eight political domains (topics): {\em External Relations, Freedom and Democracy, Political System, Economy, Welfare and Quality of Life, Fabric of Society, Social Groups} and a topic {\em undefined}, for a complete list see also \cite{leftright}. These 8 topics were used to aggregate all statements in each manifesto into topics. Most party manifestos covered all eight of them, some party manifestos in the 17th Bundestag only covered seven.

\subsection{Bag-of-Words Vectorization}\label{sec:bow-vectorization}
First each data set was segmented into semantic units; in the case of parliament discussions this were the speeches, in the case of the party manifesto data semantic units were the sentences or sentence parts associated with one of the 56 political view labels.  Parliament speeches were often interrupted; in this case each uninterrupted part of a speech was considered a semantic unit. Strings of each semantic unit were tokenised and transformed into bag-of-word vectors as implemented in scikit-learn \cite{scikit-learn}. The general idea of bag-of-words vectors is to simply count occurrences of words (or word sequences, also called {\em n-grams}) for each data point. A data point is usually a documents, here it is the semantic units of parliament speeches and manifesto sentences, respectively. The text of each semantic unit is transformed into a vector $\vec{x}\in\mathds{R}^d$ where $d$ is the size of the dictionary; the $w$th entry of $\vec{x}$ contains the (normalized) count of the $w$th word (or sequence of words) in our dictionary. We tried several options for vectorizing the speeches, including term-frequency-inverse-document-frequency normalisation, n-gram patterns up to size $n=3$ and several cutoffs for discarding too frequent and too infrequent words. All of these hyperparameters were subjected to hyperparameter optimization as explained in \autoref{sec:crossvalidation}.

\section{Classification Model and Training Procedure}\label{sec:model}
Bag-of-words feature vectors were used to train a multinomial logistic regression model. Let $y\in\{1,2,\dots,K\}$ be the true  label, where $K$ is the total number of labels and $\vec{W}=[\vec{w}_1,\dots,\vec{w}_K]\in\R^{d\times K}$ is the concatenation of the weight vectors $\vec{w}_k$ associated with the $k$th party then 
\begin{eqnarray}\label{eq:logreg_multiclass}
p(y=k|\vec{x},\vec{W}) = &\frac{e^{z_k}}{\sum_{j=1}^K e^{z_j}} \qquad \textrm{with }  z_k=&\vec{w}_k^{\top}\vec{x} \\\nonumber
\end{eqnarray}
%
We estimated $\vec{W}$ using quasi-newton gradient descent. The optimization function was obtained by adding a penalization term to the negative log-likelihood of the multinomial logistic regression objective and the optimization hence found the $\vec{W}$ that minimized
\begin{equation}\label{eq:objective}
L(\vec{W}, \vec{x}, \gamma) = - \log{\frac{e^{z_k}}{\sum_{j=1}^K e^{z_j}}}+ \gamma \| \vec{W} \|_{F}
\end{equation}
Where $\|~\|_F$ denotes the Frobenius Norm and $\gamma$ is a regularization parameter controlling the complexity of the model. 
 We optimized the regularization parameter on a log-scaled grid from $10^{-4,\dots,4}$ and the regularization constant was adopted to reflect asymmetric class frequency distributions. The performance of the model was optimized using the classification accuracy, but we also report all other standard measures, precision ($TP / (FP + TP$), recall ($TP / (TP + FN)$) and f1-score ($2\times (Prec. \times Rec) / (Prec + Rec.)$).

Three different classification problems were considered: 
\begin{enumerate}
\item {\bf Classification of party affiliation} (five class / four class problem)
\item {\bf Classification of government membership} (binary problem)
\item {\bf Classification of political views} (56 class problem)
\end{enumerate}

Party affiliation is a five class problem for the 17th legislation period, and a four class problem for the 18th legislation period. Political view classification is based on the labels of the manifesto project, see \autoref{sec:data} and \cite{leftright}. 
For each of first two problems, party affiliation and government membership prediction, we trained classifiers on the parliament speeches. For the third problem we trained only on the manifesto data for which political view labels were available.

\subsection{Optimisation of Model Parameters}\label{sec:crossvalidation}
The model pipeline contained a number of {\em hyperparameters} that were optimised using cross-validation.  
We first split the training data into a training data set that was used for optimisation of hyperparameters and an held-out test data set for evaluating how well the model performs on in-domain data; wherever possible the generalisation performance of the models was also evaluated on out-of domain data. Hyperparameters were optimised using grid search and 3-fold cross-validation within the training set only: A cross-validation split was made to obtain train/test data for the grid search and for each setting of hyperparameters the entire pipeline was trained and evaluated -- no data from the in-domain evaluation data or the out-of-domain evaluation data were used for hyperparameter optimisation. For the best setting of all hyperparameters the pipeline was trained again on all training data and evaluated on the evaluation data sets. For party affiliation prediction and government membership prediction the training and test set were 90\% and 10\%, respectively, of all data in a given legislative period. Out-of-domain evaluation data were the texts from party manifestos. For the political view prediction setting there was no out-of-domain evaluation data, so all labeled manifesto sentences in both legislative periods were split into a training and evaluation set of 90\% (train) and 10\% (evaluation).

\subsection{Sentiment analysis}\label{sec:sentiment_analysis_methods}
A publicly available key word list was used to extract sentiments from the text subjected to classification \cite{remquahey2010}. The sentiment index used for attributing positive or negative sentiment to a text was computed as the cosine similarity between bag-of-word vectors $\vec{x}\in\R^d$ and sentiment vector $\vec{s}\in\R^d$ which constructed from the values in the sentiment dictionary.

\subsection{Analysis of bag-of-words features}\label{sec:correlations_methods}
While interpretability of linear models is often propagated as one of their main advantages, doing so naively without modelling the noise covariances can lead to wrong conclusions, see e.g. \cite{Zien2009, Haufe2013}; interpreting coefficients of linear models (independent of the regularizer used) implicitly assumes uncorrelated features; this assumption is violated by the text data used in this study. Thus direct interpretation of the model coefficients $\vec{W}$ is problematic. In order to allow for better interpretation of the predictions and to assess which features contribute to the classification correlation coefficients between each word and the party affiliation label were computed. The words corresponding to the top positive and negative correlations are shown in section \autoref{sec:word_party_correlations}.

@@

The following sections give an overview of the results for all political bias prediction tasks. Some interpretations of the results are highlighted and a web application of the models is presented at the end of the section.

\subsection{Predicting political party affiliation}
The results for the political party affiliation prediction on held-out parliament data and on evaluation data are listed in \autoref{tab:results_17} for the 17th Bundestag and in \autoref{tab:results_18} for the 18th Bundestag, respectively. 
Shown are the evaluation results for in-domain data (held-out parliament speech texts) as well as the out-of-domain data results obtained from testing the models on the texts extracted from party manifestos. When predicting party affiliation on text data from the same domain that was used for training the model average precision and recall values of above 0.6 are obtained. \\

For out-of domain data the models yield significantly lower precision and recall values between 0.3 and 0.4.  
A main factor that made the prediction on the out-of-domain prediction task particularly difficult is the short length of the strings to be classified, see also \autoref{sec:data}. In order to investigate whether this low out-of-domain prediction performance was due the domain difference (parliament speech vs manifesto data) or due to the short length of the data points, we aggregated the manifesto data based on the topic. The manifesto code political topics labels were used to concatenate texts of each party to one of eight topics, see \autoref{sec:data}. The results are shown in \autoref{tab:results_topic} and \autoref{tab:confusion_topic} and demonstrate that when the texts to be classified are sufficiently long and the word count statistics are sufficiently dense the classification performance on out of domain data can achieve reliably precision and recall values close to 1.0.

@@

\begin{table}[t]
\begin{center}
\begin{tabular}{lcccc}
& \multicolumn{4}{c}{\bf 17th Bundestag} \\
    &         precision    &recall &  f1-score  & N  \\
    \hline
        \hline
cducsu     &  0.64  &    1.00  &    0.78    &     7\\
       fdp    &   1.00    &  1.00    &  1.00    &     7\\
    gruene  &     1.00  &    0.86  &    0.92    &     7\\
     linke    &   1.00   &   1.00     & 1.00    &     7\\
       spd   &    0.80   &   0.50    &  0.62     &    8\\
    \hline
    avg / total  &     0.88   &   0.86   &   0.86  &      36\\
\end{tabular}
\quad
\begin{tabular}{lcccc}
& \multicolumn{4}{c}{\bf 18th Bundestag} \\
    &         precision    &recall &  f1-score  & N  \\
    \hline
    \hline
  cducsu     &  0.50   &   1.00  &    0.67     &    8\\
       gruene  &     0.00   &   0.00   &   0.00  &       8\\
      linke     &  1.00   &   0.88   &   0.93  &       8\\
        spd     &  0.56  &    0.62  &    0.59   &      8\\
\hline
avg / total   &    0.51 &     0.62   &   0.55   &     32\\
\end{tabular}
\end{center}
\caption{
\label{tab:results_topic}
{\bf Topic level classification performance} on the party affiliation prediction problem for data from the evaluation set (manifesto texts) of the 17th legislative period. In contrast to single sentence level predictions (see \autoref{tab:results_17}, \autoref{tab:results_18}, \autoref{tab:confusion} for results and \autoref{sec:data} for topic definitions) the predictions made on topic level are reliable in many cases. Note that all manifesto topics of the green party in the 18th Bundestag are predicted to be from the parties of the governing coalition, CDU/CSU or SPD.}
\end{table}

In order to investigate the errors the models made confusion matrices were extracted for the predictions on the out-of-domain evaluation data for sentence level predictions (see \autoref{tab:confusion}) as well as topic level predictions (see \autoref{tab:confusion_topic}). One example illustrates that the mistakes the model makes can be associated with changes in the party policy. The green party has been promoting policies for renewable energy and against nuclear energy in their manifestos prior to both legislative periods. Yet the statements of the green party are more often predicted to be from the government parties than from the party that originally promoted these green ideas, reflecting the trend that these legislative periods governing parties took over policies from the green party. This effect is even more pronounced in the topic level predictions: a model trained on data from the 18th Bundestag all manifesto topics of the green party are predicted to be from one of the parties of the governing coalition, CDU/CSU or SPD. \\

\paragraph{Government membership prediction} Next to the party affiliation labels also government membership labels were used to train models that predict whether or not a text is from a party that belonged to a governing coalition of the Bundestag. In \autoref{tab:results_binary_17} and \autoref{tab:results_binary_18} the results are shown for the 17th and the 18th Bundestag, respectively. While the in-domain evaluation precision and recall values are much better in this case, reaching values close to 0.9, the out-of-domain evaluation drops again to values between 0.6 and 0.7. While topic-level predictions were not performed in this binary setting, the party affiliation results in \autoref{tab:results_topic} suggest that a similar increase in out-of-domain prediction accuracy could be achieved.

% Confusion matrix sentence level
\begin{table}[t]\label{tab:conf_mat_four_class}
\begin{tabular}{lccccccc}
 \multicolumn{8}{c}{\bf 17th Bundestag}\\
 \\
&&& \multicolumn{5}{c}{Predicted}\\
&&& cducsu & fdp& gruene& linke& spd\\
\hline
\multirow{5}{*}{\rotatebox{90}{\pbox{3cm}{\centering True}}}& &cducsu &1186 &289& 178& 198& 179\\
&&fdp &882& 658& 236& 329& 214\\
&&gruene &1174& 404& 764& 941& 464\\
&&linke &388& 92& 214& 806& 201\\
&&spd &999& 268& 240& 398& 373\\
\end{tabular}
\hfill
\begin{tabular}{lcccccc}
 \multicolumn{7}{c}{\bf 18th Bundestag}\\
 \vspace{1em}\\
&&& \multicolumn{4}{c}{Predicted}\\
&&& cducsu & gruene& linke& spd\\
\hline
\multirow{4}{*}{\rotatebox{90}{\pbox{4.7cm}{\centering True}}}&&cducsu&1912& 156& 331& 584\\
&&gruene&2092& 827& 1311& 1444\\
&&linke&596& 186& 1216& 557\\
&&spd&1284& 226& 563& 916\\
\end{tabular}
\vspace{1em}
\caption{\label{tab:confusion} {\bf Confusion matrices (sentence level)} for predictions on evaluation data (party manifestos) for classifiers trained on parliament speeches for the 17th legislative period (left) and 18th legislative period (right); the most prominent effect is the high likelihood for a party to be taken as the strongest, governing party, cdu/csu. This can be interpreted as a change in policies of the conservative party cdu/csu towards the policies of the green party.}
\end{table}

% Confusion matrix topic level
\begin{table}[t]\label{tab:conf_mat_four_class}
\begin{tabular}{lccccccc}
 \multicolumn{8}{c}{\bf 17th Bundestag}\\
 \\
&&& \multicolumn{5}{c}{Predicted}\\
&&& cducsu & fdp& gruene& linke& spd\\
\hline
\multirow{5}{*}{\rotatebox{90}{\pbox{3cm}{\centering True}}} &&cducsu &7& 0& 0& 0& 0\\
&&fdp&0& 7& 0& 0& 0\\
&&gruene&0& 0& 6& 0& 1\\
&&linke&0& 0& 0& 7& 0\\
&&spd&4& 0& 0& 0& 4\\
\end{tabular}
\hfill
\begin{tabular}{lcccccc}
 \multicolumn{7}{c}{\bf 18th Bundestag}\\
 \vspace{1em}\\
&&& \multicolumn{4}{c}{Predicted}\\
&&& cducsu & gruene& linke& spd\\
\hline
\multirow{4}{*}{\rotatebox{90}{\pbox{4.7cm}{\centering True}}}&&cducsu&8& 0& 0& 0\\
&&gruene&4& 0& 0& 4\\
&&linke&1& 0& 7& 0\\
&&spd&3& 0& 0& 5\\
\end{tabular}
\vspace{1em}
\caption{\label{tab:confusion_topic} {\bf Confusion matrices (topic level)} for predictions on evaluation data (party manifestos) for classifiers trained on parliament speeches for the 17th legislative period (left) and 18th legislative period (right).}
\end{table}
